---
title: "Analysis of Hate Crime Database"
reference: "https://raw.githubusercontent.com/fivethirtyeight/data/master/hate-crimes/hate_crimes.csv"
author: "Farhod Ibragimov"
date: "2025-01-26"
output: pdf_document
---

Loading required packages:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(ggplot2)
library(reshape2)
library(dplyr)
```

## Loading data

Let's load data from .csv file obtained from FiveThirtyEight.

Reference: ("<https://raw.githubusercontent.com/fivethirtyeight/data/master/hate-crimes/hate_crimes.csv>"):

```{r echo=TRUE}
hate_crimes <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/hate-crimes/hate_crimes.csv", header = TRUE)

```

We can see that this dataset has 51 observation points, and 12 variables.

## Correlational analysis.

Let's create correlations heatmap for numeric variables, so we can exclude state and state abbreviation variables .

```{r}
numeric_data <- hate_crimes |> 
  select_if(is.numeric)
```

```{r}
corr_hate_crimes <- cor(numeric_data, use = "complete.obs")
```

```{r}
melted_hate_crimes <- melt(corr_hate_crimes)

```

```{r}
ggplot(melted_hate_crimes, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1,1), space = "Lab", 
                       name="Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 40, vjust = 1, hjust = 1)) +
  labs(title = "Heatmap of Correlations in Hate Crimes Dataset",
       x = "Variables",
       y = "Variables")
```

Correlation values closer to 1 show positive correlation, and closer to -1 indicate negative correlation.

Values close to 0 indicate that there is no significant correlation.

Let's explore some strong positive and negative correlations

\pagebreak

## Including Plots

1.  Let's make a scatter plot to see relationship between share of white poverty and median household income

```{r}
ggplot(hate_crimes, aes(x = share_white_poverty, y = median_household_income)) +
  geom_point() +
  geom_smooth(method = "lm")+
  labs(
    title = "Share of white poverty and median household income",
    x = "share of white poverty",
    y = "median household income"
  )
```

The conclusion from this scatter plot is that increase in share of white poverty (explanatory variable) has a negative effect on median household income (response variable).

2.  Let's compare share of population in metro areas with share of non-citizens:

```{r}
ggplot(numeric_data, aes(x = share_population_in_metro_areas, y = share_non_citizen)) +
  geom_point(na.rm = TRUE)+
  geom_smooth(method = "lm")
```

In this scatter plot we can see that there is a positive relationship in share of population in metro area (explanatory) and share of non-citizens (response).

3.  Let's see how median household income related to share of voters voted for President Trump in 2016:

```{r}
ggplot(hate_crimes, aes(x = median_household_income, y = share_voters_voted_trump)) +
  geom_point(na.rm = TRUE) +
  geom_smooth(method = "lm") +
  labs(
    title = "Median household income and share of votes for President Trump in 2016",
    x = "median household income",
    y = "share of votes"
  )
```

Here we can see that increase in median household income (explanatory variable) had a negative effect on share of voters voted for President Trump (response variable) in 2016.

4.  Let's see relationship between GINI index and share of population with high school degree:

```{r}
ggplot(hate_crimes, aes(x = share_population_with_high_school_degree, y = gini_index)) +
  geom_point(na.rm = TRUE) +
  geom_smooth(method = "lm") +
  labs(
    title = " GINI index and share of population with high school degree",
    x = "share of population with high school degree",
    y = "GINI index"
  )

```

The GINI index is used to measure income inequality, wealth inequality, or consumption inequality. A GINI index of 0 represents perfect equality, while an index of 1 implies perfect inequality.

We can see that increase in share of population with high school diploma (explanatory variable) has a negative impact on GINI index (response variable).

```{r}
ggplot(hate_crimes, aes(x = median_household_income))
```
