---
title: "DATA607_WEEK9_API"
author: "Farhod Ibragimov"
date: "2025-03-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading libraries

```{r}
library(tidyverse)
library(httr)
library(jsonlite)
```

-   `httr`**:** Provides functions for working with HTTP, making it easy to send GET, POST, and other HTTP requests.

-   `jsonlite`**:** Offers tools for parsing JSON data into R objects (like lists or data frames) and converting R objects into JSON.

## Setting API URL and API key

Here I'm creating API endpoint `/svc/mostpopular/v2/viewed/1.json` for *New York Times* most viewed articles for today.

`api_key` includes API key.

```{r}
api_key <- "lucRAUjAeldYWSt6n8MQv9cGvbhwqSqr"
api_url <- paste0("https://api.nytimes.com/svc/mostpopular/v2/viewed/1.json?api-key=", api_key)
```

## Sending GET request to the NYT API

```{r}
response <- GET(api_url)
```

`GET:`this function from `httr`package sends `HTTP GET` request to the URL defined by `api_url`.

The result of the response is stored in `response` variable. It includes status information, headers, and the body of the response.

## Checking the response status

```{r}
if(status_code(response) == 200) {
  json_data <- content(response, as = "text", encoding = "UTF-8")
  data_list <- fromJSON(json_data, flatten = TRUE)
}else {
  print("Error: Unable to fetch data from API")
}
```

`status_code(response)`retrieves the HTTP status code from the response.

`200`**:** Indicates a successful request. If the status code is 200, the code proceeds to parse the content. Otherwise, it prints an error message.

## Extracting and parsing the JSON content:

```{r}
json_data <- content(response, as = "text", encoding = "UTF-8")
data_list <- fromJSON(json_data, flatten = TRUE)
print(data_list)
```

Here the code extracts the body of response as a text string.

`UTF-8`ensures that characters are correctly interpreted.

`fromJSON`converts the JSON text into `data_list`using `jsonlite.`

## Converting data into a dataframe:

```{r}
articles_df <- as.data.frame(data_list$results)
```

The API’s JSON response typically contains a key called `results` where the actual articles or items are stored.

## Displaying the dataframe:

```{r}
glimpse(articles_df)
```

```{r}
print(head(articles_df))
```
