---
title: "DATA607_WEEK9_API"
author: "Farhod Ibragimov"
date: "2025-03-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading libraries

```{r}
library(tidyverse)
library(httr)
library(jsonlite)
library(readr)
library(tidytext)
library(ggplot2)
```

-   `httr`**:** Provides functions for working with HTTP, making it easy to send GET, POST, and other HTTP requests.

-   `jsonlite`**:** Offers tools for parsing JSON data into R objects (like lists or data frames) and converting R objects into JSON.

## Setting API URL and API key

Here I'm creating API endpoint `/svc/mostpopular/v2/viewed/1.json` for *New York Times* most viewed articles for today.

`api_key` includes API key.

```{r}
api_key <- Sys.getenv("NYT_API_KEY")
api_url <- paste0("https://api.nytimes.com/svc/mostpopular/v2/viewed/30.json?api-key=", api_key)
```

## Sending GET request to the NYT API

```{r}
response <- GET(api_url)
```

`GET:`this function from `httr`package sends `HTTP GET` request to the URL defined by `api_url`.

The result of the response is stored in `response` variable. It includes status information, headers, and the body of the response.

## Checking the response status

```{r}
if(status_code(response) == 200) {
  json_data <- content(response, as = "text", encoding = "UTF-8")
  data_list <- fromJSON(json_data, flatten = TRUE)
}else {
  print("Error: Unable to fetch data from API")
}
```

`status_code(response)`retrieves the HTTP status code from the response.

`200`**:** Indicates a successful request. If the status code is 200, the code proceeds to parse the content. Otherwise, it prints an error message.

## Extracting and parsing the JSON content:

```{r}
json_data <- content(response, as = "text", encoding = "UTF-8")
data_list <- fromJSON(json_data, flatten = TRUE)
print(data_list)
```

Here the code extracts the body of response as a text string.

`UTF-8`ensures that characters are correctly interpreted.

`fromJSON`converts the JSON text into `data_list`using `jsonlite.`

## Converting data into a dataframe:

```{r}
articles_df <- as.data.frame(data_list$results)
```

The API’s JSON response typically contains a key called `results` where the actual articles or items are stored.

## Displaying the dataframe:

```{r}
glimpse(articles_df)
```

```{r}
print(head(articles_df))
```

## Analysis

This code loops through the columns of the dataframe and convert list columns to character strings.

```{r}
articles_df_clean <- data.frame(lapply(articles_df, function(col) {
  if (is.list(col)) {
    # Convert each list element to a single string (separated by "; ")
    sapply(col, function(x) paste(x, collapse = "; "))
  } else {
    col
  }
}), stringsAsFactors = FALSE)

articles_df_clean <- articles_df %>% 
  mutate(across(where(is.list), ~ sapply(.x, toString)))
```

```{r}
sections <- table(articles_df_clean$section)
sorted_section <- sort(sections, decreasing = TRUE)

sections_df <- as.data.frame(sorted_section)
colnames(sections_df) <- c("section", "count")
```

```{r}
ggplot(sections_df, aes(x = reorder(section, -count), y = count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Most viewed popular sections by count",
       x = "Section", y = "Article Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

From this plot we can say that the readers most viewed section in last 30 days is US news section.

```{r}
publish_day <- table(articles_df_clean$published_date)
sorted_publish_date <- sort(publish_day, decreasing = TRUE)
date_df <- as.data.frame(sorted_publish_date)
colnames(date_df) <- c("published_date", "count")
```

```{r}

ggplot(date_df, aes(x = reorder(published_date, -count), y = count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Number of Popular Articles by Date",
       x = "Published Date", y = "Article Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

From this chart we can see that most popular articles views day in last 30 days is March 7th, 2025
