---
title: "Tidying US Census dataset"
author: "Farhod Ibragimov"
date: "2025-03-04"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(usmap)
library(dplyr)

```

## Tidying dataset

```{r}
census_data <- read.csv("https://raw.githubusercontent.com/farhodibr/CUNY-SPS-MSDS/refs/heads/main/DATA607/PROJECT2/US_Census_Data.csv")

census_data <- census_data |>
  select(-Fact.Note)
glimpse(census_data)
```

Why do I think this dataset is not tidy:

-   **Multiple variables in one column**: \
    "Fact" column contains several descriptions of the data, which needs to be separate to be tidy

-   States supposed to be rows (observations), not columns.

-   Each row represents multiple observations. In tidy dataset each row represents single observation, for example a specific state's demographic data for a single year.

Here I create separate tidy data table for populations of each state in years 2010 and 2016

```{r}

state_names <- colnames(census_data)[3:ncol(census_data)]
#state_names

population_data <- census_data |>
  filter(Fact %in% c("Population estimates, July 1, 2016,  (V2016)", 
                     "Population, Census, April 1, 2010"))

long_population_data <- population_data |>
  pivot_longer(
    cols = -Fact,
    names_to = "state",
    values_to = "population"
  ) |>
  mutate(year = case_when(
    grepl("2016", Fact) ~2016,
    grepl("2010", Fact) ~ 2010,
    TRUE ~ NA_integer_
  )) |>
  select(-Fact) |>
  pivot_wider(
    names_from = year,
    values_from = population
   ) |>
  # |>
  # mutate(
  #   `2010` = as.numeric(`2010`),
  #   `2016` = as.numeric(`2016`)
  # )
  select(state, "2010", "2016")

long_population_data$`2010` <- gsub(",", "", long_population_data$`2010`)
long_population_data$`2010` <-  as.numeric(long_population_data$`2010`)

long_population_data$`2016` <- gsub(",", "", long_population_data$`2016`)
long_population_data$`2016` <-  as.numeric(long_population_data$`2016`)
print(long_population_data)
```

This `long_population_data`data table is tidy and ready for analysis.

In this code I created `create_long_table`function which makes it easier to create different tidy data tables:

```{r}
create_long_table <- function(column_name, rows) {
  result <- census_data |>
  slice(rows) |>
  pivot_longer(
    cols = !Fact,
    names_to = "state",
    values_to = column_name
  ) |>
   mutate(year = case_when(
    grepl("2016", Fact) ~2016,
    grepl("2010", Fact) ~ 2010,
    TRUE ~ NA_integer_
  )) |>
  select(-Fact) |>
  select(state, year, all_of(column_name)) |>
  mutate(
    !!column_name := round(as.numeric(gsub("%", "", !!sym(column_name))), 2)
    )
  
  return(result)
}


```

Here I create `gender_table_long`tidy data table which includes female population proportions for each state in years 2010 and 2016. I use `create_long_table`to create this table. Also I did data transformation because some states had proportions in percents, and some decimal values as actual proportions.

```{r}
gender_table_long <- create_long_table("female_prop", 11:12) 
gender_table_long <- gender_table_long |>
  mutate(
    female_prop = if_else(
                          female_prop < 1,
                          round(female_prop * 100, 2),
                          round(female_prop, 2)
  ))
print(gender_table_long)



```

This code creates new observations in `gender_table_long` for male population for each state which is better for analysis.

```{r}
gender_table_long|>
  mutate(
    male_prop = 100 - female_prop
  ) |>
  pivot_longer(
    cols = contains("prop"),
    names_to = "gender",
    values_to = "value"
  ) |>
  pivot_wider(
    names_from = year,
    values_from = value,
    names_prefix = "X"
  ) |>
  select(state, gender, X2010, X2016) |>
  mutate(
    prop_change = X2016 - X2010
  )
head(gender_table_long)
```

From here I created few data tables for different age ranges:

```{r}
under_5_proportions_long <- create_long_table("prop_under_5", 5:6) |>
  group_by(year, state)
print(under_5_proportions_long)
```

```{r}
under_18_proportions_long <- create_long_table("prop_under_18", 7:8)
print(under_18_proportions_long)
```

```{r}
over_65_proportions_long <- create_long_table("over_65", 9:10)
print(over_65_proportions_long)
```

```{r}
population_proportions <- under_5_proportions_long |>
  left_join(under_18_proportions_long, by = c("state", "year")) |>
  left_join(over_65_proportions_long, c("state", "year")) |>
  mutate(
    prop_18_to_65 = 100 - prop_under_18 - over_65
  )
```

This code creates data table for population proportions in 18-65 ages range for each state

```{r}
population_18_65_long <- population_proportions |>
  select(state, year, prop_18_to_65)
print(population_18_65_long)
```

```{r}



```

## Analysis

This plot shows analysis for each state's population in 2010.

```{r}
library(usmap)
long_population_data$state <- gsub("\\.", " ", long_population_data$state) 

names(long_population_data)[names(long_population_data) == "2010"] <- "pop_2010"
names(long_population_data)[names(long_population_data) == "2016"] <- "pop_2016"

data_for_map <- long_population_data |>
  left_join(usmap::statepop, by = c("state" = "full"))

mismatches <- setdiff(long_population_data$state, usmap::statepop$full)
print(mismatches)

# long_population_data$state <- trimws(long_population_data$state)
# usmap::statepop$full <- trimws(usmap::statepop$full)
# head(data_for_map)
# plot_usmap(data_for_map, values = "pop_2010", color = "blue") +
#   scale_fill_continuous(name = "Population 2010", 
#                         low = "white",
#                         high = "blue") +
#   theme(legend.position = "right") +
#   labs(title = "US State Population in 2010")

plot_usmap(data = long_population_data, values = "pop_2010", color = "blue") +
  scale_fill_continuous(name = "Population 2010", 
                        low = "white", 
                        high = "blue",
                        labels = scales::comma) +
  theme(legend.position = "right") +
  labs(title = "US State Population in 2010")
```

As we see on the plot most populated states in 2010:

1.  California

2.  Texas

3.  Florida

4.  New York\
    \
    This plot shows how population changed in states in period of 2010 - 2016\

```{r}

long_population_data <- long_population_data |>
  mutate(
    pop_change = pop_2016 - pop_2010
  )

plot_usmap(data=long_population_data, values = "pop_change", color = "gray") +
  scale_fill_gradient2(name = "Population change 2010 - 2016",
                        low = "red", mid = "blue", high = "green",
                        midpoint = 0,
                        labels = scales::comma
                        ) +
  theme(legend.position = "right") +
  labs(title = "Population Chanhe in US State Population 2010 - 2016") 
```

From this plot we see states which had most increases in population:

1.  Texas

2.  California

3.  Florida
