---
title: "Cola"
author: "Farhod Ibragimov"
date: "2025-03-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Tidying "Cola" dataset

```{r}
library(tidyverse)
library(readr)
```

```{r}
data <- read.csv("https://raw.githubusercontent.com/farhodibr/CUNY-SPS-MSDS/refs/heads/main/DATA607/PROJECT2/Cola.csv")
head(data)
```

Why is this dataset untidy:

-   Multiple variables in separate rows: The dataset contains various variables that are scattered across rows.

-   Fiscal years as columns

This dataset is big and could use a lot of tidying and broken down into separate tidy data tables. I will just use Profit & Loss statement to create tidy data table as example.

```{r}
profit_data <- data |>
  slice(4:6) |>
  select(!Data.provided.by.SimFin)
print(profit_data)
```

Here I am cleaning and formating rows and column names:

```{r}
colnames(profit_data) <- tolower(gsub(" ", "_", colnames(profit_data)))
profit_data$x <- tolower(gsub(" ", "_", profit_data$x))

profit_data <- profit_data |>
  mutate(across(where(is.character), ~ tolower(.)))

profit_data <- profit_data |>
  rename( profit_loss = x)
         
print(profit_data)
```

This code first create a long table, and after I am making it wide to make it tidy:

```{r}
profit_data_tidy <- profit_data |>
  pivot_longer(
    col = starts_with("x"),
    names_to = "period",
    values_to = "amount"
  ) |>
  group_by(profit_loss, period)
head(profit_data_tidy)
profit_data_tidy$amount <- as.numeric(gsub(",", "", profit_data_tidy$amount))
profit_data_tidy$period <- as.numeric(gsub("x.", "", profit_data_tidy$period)) + 2000

profit_data_tidy <- profit_data_tidy |>
  pivot_wider(
    names_from = profit_loss,
    values_from = amount
  )
head(profit_data_tidy)
```

Why is this dataset is tidy?

-   Each row represents a single observation which corresponds to a specific period (year 2001, 2002, etc.), making it easy to analyze trends or patterns over time.

-   The dataset has separate columns for `period`, `net_operating_revenues`, `cost_of_goods_sold`, and `gross_profit`, which allows each variable to be analyzed independently.

## Analysis

Here is the plot showing financial performance over time:

```{r}
#profit_data_tidy$period <- as.factor(profit_data_tidy$period)

profit_data_tidy$net_operating_revenues <- as.numeric(profit_data_tidy$net_operating_revenues)
profit_data_tidy$cost_of_goods_sold <- as.numeric(profit_data_tidy$cost_of_goods_sold)
profit_data_tidy$gross_profit <- as.numeric(profit_data_tidy$gross_profit)

ggplot(profit_data_tidy, aes(x = period)) + 
  geom_line(aes(y = net_operating_revenues, 
                color = "Net Operating Revenues", group = 1), 
                linewidth = 1) + 
  geom_line(aes(y = cost_of_goods_sold, 
                color = "Cost of Goods Sold", group = 1), 
                linewidth = 1) +
  geom_line(aes(y = gross_profit, 
                color = "Gross Profit", group = 1), 
                linewidth = 1) +
  labs(title = "Trend Analysis: Financial Performance Over Time",
       x = "Period", 
       y = "Amount (in USD)",
       color = "Legend") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

From this plot I can say:

-   Revenue increased from 2001 to 2004. From 2005 it started to decline and in 2010 revenue was approximately same as in 2001.

This plot analyze profit margin over the years

```{r}
profit_data_tidy <- profit_data_tidy |>
  mutate(gross_profit_margin = (gross_profit / net_operating_revenues) * 100)

profit_data_tidy$period <- as.numeric(as.character(profit_data_tidy$period))

print(profit_data_tidy)

ggplot(profit_data_tidy, aes(x = period, y = gross_profit_margin)) +
  geom_point(color = "red") + 
  geom_line(color = "blue", linewidth = 1) +
  labs(title = "Gross Profit Margin Over Time",
       x = "Period", y = "Gross Profit Margin (%)") +
  scale_x_continuous(breaks = seq(min(profit_data_tidy$period), max(profit_data_tidy$period), by = 1),
                     labels = as.character(seq(min(profit_data_tidy$period), max(profit_data_tidy$period), by = 1))) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Profit margin declined significantly in years of 2002-2004, which is interesting. Because from the previous plot we can see that revenues peak is in 2004. As we can see from 2007 profit margin steadily increasing.
