---
title: "Project_2_DATA_607 Emissions dataset tidying"
author: "Farhod Ibragimov"
date: "2025-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(tidyr)
library(dplyr)
library(readr)
library(data.table)
library(maps)
library(sf)
```

## Tidying and cleaning data

Why do I think this dataset is not tidy:

-   Multiple variables in separate columns. Each row should represent a single observation (e.g., emissions for a specific region, item, and year), but this is not the case here.

-   Years as columns. \
    \
    This code loads dataset, also it cleans and transform data:

```{r}
emissions_data <- read.csv("https://raw.githubusercontent.com/farhodibr/CUNY-SPS-MSDS/refs/heads/main/DATA607/PROJECT2/Total%20Emissions%20Per%20Country%20(2000-2020)%20(1).csv")

head(emissions_data)


year_cols <- grep("^X2", colnames(emissions_data), value = TRUE)
emissions_data <- emissions_data |>
  distinct(across(all_of(starts_with("X2"))), .keep_all = TRUE)


format_element <- function(x) {
  letters_match <- str_extract(x, "(?i)(CO2eq|N2O|CH4|CO2)")
  clean_txt <- str_remove(x, "(?i)(CO2eq|N2O|CH4|CO2)\\s*") |>
    str_remove_all("(?i)\\bemissions\\b")
  
  clean_txt <- str_to_lower(clean_txt) |>
    str_replace_all("\\s+", "_") |>
    str_replace_all("[^a-z0-9_]", "")|>
    str_replace_all("_$", "")
  
  result <- if(!is.na(letters_match) &letters_match !="") {
    str_c(letters_match, clean_txt, sep = "_")
  } else {
    clean_txt
  }
  
  result <- str_replace_all(result, "(?i)\\bn2o\\b", "N2O")
  result <- str_replace_all(result, "(?i)\\bch4\\b", "CH4")
  result <- str_replace_all(result, "(?i)\\bar5\\b", "AR5")
  result <- str_replace_all(result, "(?i)\\bco2\\b", "CO2")
  result <- str_replace_all(result, "_$", "")
  
  return(result)
}

colnames(emissions_data) <- str_remove_all(colnames(emissions_data), "(?i)^X")

emissions_data <- emissions_data |>
  mutate(Element = map_chr(Element, format_element)) |>
  drop_na()
head(emissions_data)
```

```{r}
split_row <- 31406

emissions_data_by_countries <- emissions_data[1:split_row, ]
emissions_data_by_continents <- emissions_data[(split_row +1):nrow(emissions_data), ]
```

Here I am creating tidy data table:

```{r}
emissions_data_long <- emissions_data_by_countries |>
  
  pivot_longer(
    cols = starts_with("2"),
    names_to = "Year",
    values_to = "Emission"
  ) |>
  drop_na()
colnames(emissions_data_long) <- str_to_lower(colnames(emissions_data_long))
emissions_data_long$year <- as.numeric(emissions_data_long$year)
emissions_data_long$emission <- round(emissions_data_long$emission, 2)

unique_areas <- unique(emissions_data_long$area)
#print(unique_areas)
#write.csv(emissions_data_long, "emissions_data_long.csv", row.names = FALSE)
head(emissions_data_long)


```

This dataset is tidy because:

-   Each variable has its own column. The columns represent individual variables: `area`, `item`, `element`, `unit`, `year`, and `emission`. Each column holds one type of data.

-   Each observation has its own row. This makes it easier for analysis.

## Analysis

Here I do analysis for top 10 countries for volumes of emission they produce:

```{r}
#total emissions by country
total_emissions_by_country <- emissions_data_long |>
  group_by(area) |>
  summarise(total_emission = sum(emission, na.rm = TRUE)) |>
  arrange(desc(total_emission))

top10_emissions <- total_emissions_by_country |>
  top_n(10, total_emission)

ggplot(top10_emissions, aes(x = reorder(area, total_emission), 
                                       y = total_emission)) +
  geom_bar(stat = "identity", fill = "steelblue")+
  scale_y_continuous(trans = "log10")+
  coord_flip() +
  labs(title = "Top Total Emissions by Country 2000 - 2020",
       x= "Country",
       y = "Total Emissions (Kilotones)")+
  theme_minimal()+
  theme(axis.text.y = element_text(angle = 45, hjust = 1))
```

In this plot we see top 10 countries with highest amounts of emissions produced in years 2000 - 2020.
