---
title: "DATA 606 Data Project Proposal"
author: "Farhod Ibragimov"
output: pdf_document
---

### Data Preparation

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
# load data
bank_data <- read.csv("https://raw.githubusercontent.com/farhodibr/CUNY-SPS-MSDS/refs/heads/main/DATA606/PROJECT/PROPOSAL/bank.csv", sep = ";")
?bank_data
```

```{r}
library(tidyverse)
library(ggplot2)
```

```{r}
head(bank_data)
```

```{r}
glimpse(bank_data)
```

### Research question

In this research project i want to see if we can predict if a client will subscribe to a term deposit (which is variable `'y'`) , based on factors such as age, job, marital status, education, and call duration.

### Cases

Each observation in database is a marketing call to the client. There are 4521 observations in this dataset.

### Data collection

The data is related with direct marketing campaigns (phone calls) of a Portuguese banking institution. 

### Type of study

This is an observational study.

### Data Source

This dataset is available at UCI machine learning repository

### Describe your variables?

There are total of 17 variables in this dataset. The table below shows type and description for each variable.

| Variable | Type | Description |
|-------------|-------------|---------------------------------------------|
| age | Quantitative | Age of the client |
| job | Qualitative | Type of job (e.g., admin., technician, etc.) |
| marital | Qualitative | Marital status (e.g., married, single, divorced) |
| education | Qualitative | Level of education (e.g., primary, secondary, tertiary) |
| default | Qualitative | Has credit in default? (yes/no) |
| balance | Quantitative | Average yearly balance in euros |
| housing | Qualitative | Has housing loan? (yes/no) |
| loan | Qualitative | Has personal loan? (yes/no) |
| contact | Qualitative | Contact communication type (e.g., cellular, telephone) |
| day | Quantitative | Last contact day of the month |
| month | Qualitative | Last contact month of the year |
| duration | Quantitative | Last contact duration in seconds |
| campaign | Quantitative | Number of contacts during this campaign for this client |
| pdays | Quantitative | Days since the client was last contacted (999 = never) |
| previous | Quantitative | Number of contacts before this campaign |
| poutcome | Qualitative | Outcome of the previous marketing campaign |
| y | Qualitative | Has the client subscribed to a term deposit? (yes/no) (target var) |

The dependent variable is `'y',`which is qualitive and indicatesif the client subscribed to a term deposit? (yes/no).

All other variables will be tried and used as independent variables in this project.

### Relevant summary statistics

**Provide summary statistics for each the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r}
summary(bank_data)
```

### Plots.

```{r}
ggplot(bank_data, aes(x = age)) +
  geom_histogram(fill = "orange", color = "blue") +
  labs(title = "Distribution of clients age")
```

The distribution of clients age appears as close to normal, right-skewed meaning there more younger clients than older ones. Most clients age is in the range of 25 to 45 years old, with the highest mode of 32-33 years old.

```{r}
ggplot(bank_data, aes(x = balance)) +
  geom_histogram(binwidth = 1000,  fill = "orange", color = "blue") +
  labs(title = "Distribution of clients balances")
```

The distribution of balances is right-skewed. I can say that many clients have low and even negative balances. There few outliers, which indicate that few clients have extremely high balances.

```{r}
ggplot(bank_data, aes(x = duration)) +
  geom_histogram(binwidth = 50, fill = "orange", color = "blue") +
  labs(title = "Distribution of call durations in seconds",
       x = "Call duration in seconds",
       y = "Count")
```

Distribution of call durations in seconds is right-skewed with a long tail of longer calls. Most of calls duration is in the range of 50 - 500 seconds. There are few calls with extremely high durations

```{r}
ggplot(bank_data, aes(x = job)) +
  geom_bar(fill = "orange") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribution of job types")
```

From this plot we can see that majority of client are employed in blue-collar, management and technician types of job.

```{r}
ggplot(bank_data, aes(x = y)) +
  geom_bar(fill = "orange") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribution of current campaign outcomes",
       x = "Outcome")
```

From this plot I can say that current campaign is not successfull, with very small amount of clients subscribed to a term deposit.

### **Research question**

I will use Logistics Regression to model if the customer subscribes to a term deposit (Variable y = "yes"). I will use the rest of variables as predictors.
