---
title: "The normal distribution"
author: "Farhod Ibragimov"
output:
  pdf_document: default
  html_document:
    includes:
      in_header: header.html
    css: ./lab.css
    highlight: pygments
    theme: cerulean
    toc: true
    toc_float: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r echo = FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE)
```

In this lab, you'll investigate the probability distribution that is
most central to statistics: the normal distribution. If you are
confident that your data are nearly normal, that opens the door to many
powerful statistical methods. Here we'll use the graphical tools of R to
assess the normality of our data and also learn how to generate random
numbers from a normal distribution.

## Getting Started

### Load packages

In this lab, we will explore and visualize the data using the
**tidyverse** suite of packages as well as the **openintro** package.

Let's load the packages.

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(patchwork)
```

### The data

This week you'll be working with fast food data. This data set contains
data on 515 menu items from some of the most popular fast food
restaurants worldwide. Let's take a quick peek at the first few rows of
the data.

Either you can use `glimpse` like before, or `head` to do this.

```{r load-data, results=TRUE}
library(tidyverse)
library(openintro)
data("fastfood", package='openintro')
head(fastfood)
```

You'll see that for every observation there are 17 measurements, many of
which are nutritional facts.

You'll be focusing on just three columns to get started: restaurant,
calories, calories from fat.

Let's first focus on just products from McDonalds and Dairy Queen.

```{r}
mcdonalds <- fastfood %>%
  filter(restaurant == "Mcdonalds")
dairy_queen <- fastfood %>%
  filter(restaurant == "Dairy Queen")
summary(mcdonalds)
```

1.  Make a plot (or plots) to visualize the distributions of the amount
    of calories from fat of the options from these two restaurants. How
    do their centers, shapes, and spreads compare?

**Insert your answer here**

*Farhod's answer:*

Let's make histogram plots for each distribution :

```{r}
mean_calfat_mcd <- mean(mcdonalds$cal_fat, na.rm = TRUE)
median_calfat_mcd <- median(mcdonalds$cal_fat, na.rm = TRUE)

ggplot(mcdonalds, aes(x = cal_fat)) +
  geom_histogram(fill = "palegreen") +  
  
  geom_vline(aes(xintercept = mean_calfat_mcd), 
             size = .8, color = "blue", 
             linetype = "dashed") +
   geom_vline(aes(xintercept = median_calfat_mcd), 
              size = .8, color = "red", 
              linetype = "dashed") +
  annotate("text", x = mean_calfat_mcd, y = 0.002, 
           label = paste("Mean (blue line) = ", round(mean_calfat_mcd, 2)), 
           vjust = -15, hjust = -0.1, color = "blue") +
  annotate("text", x = median_calfat_mcd, y = 0.002, 
           label = paste("Median (red line) = ", round(median_calfat_mcd, 2)), 
           vjust = -17, hjust = -0.22, color = "red") +
  labs(title = "Distribution of calories from fat in McDonals's",
       x = "Calories from fat",
       y = "Counts") +
  theme_minimal() 
```

```{r}
mean_dq_clafat <- mean(dairy_queen$cal_fat, na.rm = TRUE)
median_dq_calfat <- median(dairy_queen$cal_fat, na.rm = TRUE)

ggplot(dairy_queen, aes(x = cal_fat)) +
  geom_histogram(  fill = "palegreen") +  
  
  geom_vline(aes(xintercept = mean_dq_clafat), 
             size = .8, color = "blue", 
             linetype = "dashed") +
   geom_vline(aes(xintercept = median_dq_calfat), 
              size = .8, color = "red", 
              linetype = "dashed") +
  annotate("text", x = mean_dq_clafat, y = 0.002, 
           label = paste("Mean (blue line) = ", round(mean_dq_clafat, 2)), 
           vjust = -20, hjust = -0.1, color = "blue") +
  annotate("text", x = median_dq_calfat, y = 0.002, 
           label = paste("Median (red line) = ", round(median_dq_calfat, 2)), 
           vjust = -22, hjust = -0.22, color = "red") +
  labs(title = "Distribution of calories from fat in Dairy Queens",
      x = "Calories from fat", 
      y = "Counts"
       ) +
  theme_minimal() 

```

Here is comparisons of centers, shapes, and spreads of distributions.

-   Centers:\
    Distribution of Dairy Queen has the mean of 260.48 and the median of
    220 calories.\
    Distribution of McDonald's calories from fat has the mean of 285.61
    and the median of 240 calories.\
    Because both distributions means are larger then their medians, this
    indicates that distributions are right (positively) skewed. We can
    see that from the plots as well.

-   Shapes:\
    Distribution of calories from fat in Dairy Queens items is
    multimodal indicating that there are three clusters of items with
    similar calories from fat counts. It has a larger cluster of data in
    a range from 80 to 350 calories..\
    Distribution of McDonald's data of calories from fat has two modes
    and is bimodal with the large cluster of data in the range from 50
    to 470 calories\
    Both shapes has longer tails to the right and that suggests
    distributions are deviating from the normal distribution.

-   Spreads:\
    McDonald's data spread is in range from 50 to 1270 calories from
    fat.\
    Data spread of calories from fat in Dairy Queens items ranges from 0
    to 670 calories.\
    McDonald's has a larger spread of distribution to compare with Dairy
    Queens spread of distribution.

*End of answer.*

## The normal distribution

In your description of the distributions, did you use words like
*bell-shaped*or *normal*? It's tempting to say so when faced with a
unimodal symmetric distribution.

To see how accurate that description is, you can plot a normal
distribution curve on top of a histogram to see how closely the data
follow a normal distribution. This normal curve should have the same
mean and standard deviation as the data. You'll be focusing on calories
from fat from Dairy Queen products, so let's store them as a separate
object and then calculate some statistics that will be referenced later.

```{r}
dqmean <- mean(dairy_queen$cal_fat)
dqsd   <- sd(dairy_queen$cal_fat)
```

Next, you make a density histogram to use as the backdrop and use the
`lines` function to overlay a normal probability curve. The difference
between a frequency histogram and a density histogram is that while in a
frequency histogram the *heights* of the bars add up to the total number
of observations, in a density histogram the *areas* of the bars add up
to 1. The area of each bar can be calculated as simply the height
*times* the width of the bar. Using a density histogram allows us to
properly overlay a normal distribution curve over the histogram since
the curve is a normal probability density function that also has area
under the curve of 1. Frequency and density histograms both display the
same exact shape; they only differ in their y-axis. You can verify this
by comparing the frequency histogram you constructed earlier and the
density histogram created by the commands below.

```{r}
ggplot(data = dairy_queen, aes(x = cal_fat)) +
        geom_blank() +
        geom_histogram(aes(y = ..density..)) +
        stat_function(fun = dnorm, args = c(mean = dqmean, sd = dqsd), col = "tomato")
```

After initializing a blank plot with `geom_blank()`, the `ggplot2`
package (within the `tidyverse`) allows us to add additional layers. The
first layer is a density histogram. The second layer is a statistical
function -- the density of the normal curve, `dnorm`. We specify that we
want the curve to have the same mean and standard deviation as the
column of fat calories. The argument `col` simply sets the color for the
line to be drawn. If we left it out, the line would be drawn in black.

2.  Based on the this plot, does it appear that the data follow a nearly
    normal distribution?

**Insert your answer here**

*Farhod's answer:*

Histogram has a multiple modes, and bins of actual distribution are much
higher than simulated normal distribution curve. Both histogram and the
line have longer tails to the right suggesting the distribution is
right-skewed. These findings can suggest that distribution is not near
normal distribution.

However, when I imagine a curve over the shape of actual distribution,
the major part of it appears similary close to simulated normal
distribution bell-shaped curve.

Still there is not enough evidence to build enough confidence to say
that distribution of calories from fat in Dairy Queens items is near
normal distribution.

*End of answer.*

## Evaluating the normal distribution

Eyeballing the shape of the histogram is one way to determine if the
data appear to be nearly normally distributed, but it can be frustrating
to decide just how close the histogram is to the curve. An alternative
approach involves constructing a normal probability plot, also called a
normal Q-Q plot for "quantile-quantile".

```{r}
ggplot(data = dairy_queen, aes(sample = cal_fat)) + 
  geom_line(stat = "qq")
```

This time, you can use the `geom_line()` layer, while specifying that
you will be creating a Q-Q plot with the `stat` argument. It's important
to note that here, instead of using `x` instead `aes()`, you need to use
`sample`.

The x-axis values correspond to the quantiles of a theoretically normal
curve with mean 0 and standard deviation 1 (i.e., the standard normal
distribution). The y-axis values correspond to the quantiles of the
original unstandardized sample data. However, even if we were to
standardize the sample data values, the Q-Q plot would look identical. A
data set that is nearly normal will result in a probability plot where
the points closely follow a diagonal line. Any deviations from normality
leads to deviations of these points from that line.

The plot for Dairy Queen's calories from fat shows points that tend to
follow the line but with some errant points towards the upper tail.
You're left with the same problem that we encountered with the histogram
above: how close is close enough?

A useful way to address this question is to rephrase it as: what do
probability plots look like for data that I *know* came from a normal
distribution? We can answer this by simulating data from a normal
distribution using `rnorm`.

```{r}
set.seed(123)
sim_norm <- rnorm(n = nrow(dairy_queen), mean = dqmean, sd = dqsd)
sim_norm
```

The first argument indicates how many numbers you'd like to generate,
which we specify to be the same number of menu items in the
`dairy_queen` data set using the `nrow()` function. The last two
arguments determine the mean and standard deviation of the normal
distribution from which the simulated sample will be generated. You can
take a look at the shape of our simulated data set, `sim_norm`, as well
as its normal probability plot.

3.  Make a normal probability plot of `sim_norm`. Do all of the points
    fall on the line? How does this plot compare to the probability plot
    for the real data? (Since `sim_norm` is not a data frame, it can be
    put directly into the `sample` argument and the `data` argument can
    be dropped.)

**Insert your answer here**

*Farhod's answer:*

```{r}
ggplot(, aes(sample = sim_norm)) +
   geom_line(stat = "qq")

```

The plot above surprisingly has slight deviations from straight diagonal
line with especially heavy deviations at both tails. Heavy deviations at
tails suggest that simulated data from a normal distribution has more
extreme values (outliers). It is not typicaly expected for true normal
distribution to have outliers located further away from the mean.

While comparing Q-Q plot for Dairy Queen data with the Q-Q plot of
simulated normal probability plot I see that both Q-Q plots are
following close to straight diagonal line. Each plot has deviations from
the straight diagonal line, especially with heavier deviations at the
tails.

However plot drawn for Dairy Queen data has more significant deviations
comparing with deviations of simulated normal probability plot. I also
noticed that the line in Dairy Queens plot has a significant jump close
to theoretical standard deviation (1), and it suggests that Dairy Queens
data has more extremely heavier outliers.

*End of answer.*

Even better than comparing the original plot to a single plot generated
from a normal distribution is to compare it to many more plots using the
following function. It shows the Q-Q plot corresponding to the original
data in the top left corner, and the Q-Q plots of 8 different simulated
normal data. It may be helpful to click the zoom button in the plot
window.

```{r}
set.seed(134)
qqnormsim(sample = cal_fat, data = dairy_queen)
```

4.  Does the normal probability plot for the calories from fat look
    similar to the plots created for the simulated data? That is, do the
    plots provide evidence that the calories are nearly normal?

**Insert your answer here**

*Farhod's answer:*

From these plots I can see overall similarity in actual data plot and
plots of 8 different simulated normal data. They have similar deviations
from straight diagonal line expected in normal distribution. Especially
there are similar deviations in right tails in data, sim 2 and sim 3
plots.

I can say that these plots show strong evidence suggesting data of
calories from fat of Dairy Queens menu items is nearly normal
distributed.

*End of answer.*

5.  Using the same technique, determine whether or not the calories from
    McDonald's menu appear to come from a normal distribution.

**Insert your answer here**

*Farhod's answer:*

```{r}
set.seed(138)
qqnormsim(sample = cal_fat, data = mcdonalds)
```

Data plot show significant jump of the right tail forming a curved line,
which is significant deviation from expected straight line of normal
distribution. Also I don't see similarity between plot of data and plots
created for the simulated data. This suggests that McDonald's data is
not near normal distribution.

*End of answer.*

## Normal probabilities

Okay, so now you have a slew of tools to judge whether or not a variable
is normally distributed. Why should you care?

It turns out that statisticians know a lot about the normal
distribution. Once you decide that a random variable is approximately
normal, you can answer all sorts of questions about that variable
related to probability. Take, for example, the question of, "What is the
probability that a randomly chosen Dairy Queen product has more than 600
calories from fat?"

If we assume that the calories from fat from Dairy Queen's menu are
normally distributed (a very close approximation is also okay), we can
find this probability by calculating a Z score and consulting a Z table
(also called a normal probability table). In R, this is done in one step
with the function `pnorm()`.

```{r}
1 - pnorm(q = 600, mean = dqmean, sd = dqsd)
```

Note that the function `pnorm()` gives the area under the normal curve
below a given value, `q`, with a given mean and standard deviation.
Since we're interested in the probability that a Dairy Queen item has
more than 600 calories from fat, we have to take one minus that
probability.

Assuming a normal distribution has allowed us to calculate a theoretical
probability. If we want to calculate the probability empirically, we
simply need to determine how many observations fall above 600 then
divide this number by the total sample size.

```{r}
dairy_queen %>% 
  filter(cal_fat > 600) %>%
  summarise(percent = n() / nrow(dairy_queen))
```

Although the probabilities are not exactly the same, they are reasonably
close. The closer that your distribution is to being normal, the more
accurate the theoretical probabilities will be.

6.  Write out two probability questions that you would like to answer
    about any of the restaurants in this dataset. Calculate those
    probabilities using both the theoretical normal distribution as well
    as the empirical distribution (four probabilities in all). Which one
    had a closer agreement between the two methods?

**Insert your answer here**

*Farhod's answer:*

**Question 1: What is the probability that a randomly chosen Dairy Queen
item has at least 300 and less than 500 calories from fat?**

To answer this question let's use the following methods.

-   Assuming normal distribution:\
    \
    This method assumes that calories from fat in Dairy Queens items are
    normally distributed.\
    First let's find theoretical probabilities of items with equal or
    less than 300 :

```{r}
prob_atleast_300 <- pnorm(q = 300, mean = dqmean, sd = dqsd)

prob_atleast_300
```

Now let's find out what are theoretical probabilities for items with
less than 500 calories:

```{r}
prob_more_than_500 <- pnorm(q = 500, mean = dqmean, sd = dqsd)

prob_more_than_500
```

To find the probability of value falling between 300 and 500 calories,
we need to subtract the probability of 300 calories from the probability
of 500 calories:

```{r}
prob_range_300_500 <- prob_more_than_500 - prob_atleast_300
prob_range_300_500
```

The theoretical probability of value falling between 300 and 500
calories is equal to 0.3374. That means that 33.74% of calories from fat
in assumed normal distribution data has values falling between 300 and
500 calories.

-   Empirical calculation:\
    \
    We need to filter actual data to keep values of calories from fat
    that at least 300 and less than 500 calories. Dividing count of rows
    in filtered data by the by the count of all rows of actual data
    gives the empirical probability for values falling between 300 and
    500 calories:

```{r}
empirical_prob <- dairy_queen |>
  filter(cal_fat >= 300 & cal_fat < 500) |>
  nrow() / nrow(dairy_queen)

empirical_prob
```

The empirically calculated probability is 0.2619, which mean that 26.19%
of calories from fat data in Dairy Queens has values falling between 300
and 500 calories.

**Question 2: What is the probability that a randomly chosen Dairy Queen
item has less than 400 calories from fat?**

-   Assuming normal distribution:

    ```{r}
    prob_lessthan_400_theoretical <- pnorm(q = 400, mean = dqmean, sd = dqsd)
    prob_lessthan_400_theoretical

    ```

    The theoretical probability is 0.8137, which means 81.37% chance
    that randomly chosen item in the assumed normally distributed data
    will have less than 400 calories.

-   Empirical calculation:\

    ```{r}
    empirical_prob <- dairy_queen |>
      filter(cal_fat < 400) |>
      nrow() / nrow(dairy_queen)
    empirical_prob
    ```

    This calculation shows that 78.57% chance that randomly chosen item
    in the actual calories from fat in Dairy Queens data will have less
    than 400 calories. This is very close to theoretically calculated
    81.37% probability chance.

Conclusion: \
In the Question 1 theoretical (33.74%) is more higher than empirical
(26.19%) calculation with the difference of 33.74% - 26.19% = 7.55%.
This higher difference suggests wider intervals in the range. This is
because actual data is right skewed as we saw in the histogram before
and the majority of the data clustered below the value of 300. That is
the reason for empirical calculation is lower than theoretical.

The Question 2 has very close theoretical (81.37%) and empirical
(78.57%) calculations with the difference 81.37% - 78.57% = 2.8%.

This findings conclude that Question 2 has a closer agreement between
the two methods.

------------------------------------------------------------------------

## More Practice

7.  Now let's consider some of the other variables in the dataset. Out
    of all the different restaurants, which ones' distribution is the
    closest to normal for sodium?

**Insert your answer here**

*Farhod's answer:*

```{r}
sodium_group <- fastfood |>
  group_by(restaurant) |>
  select(restaurant, sodium)

ggplot(sodium_group, aes(sample = sodium)) +
  stat_qq() +
  stat_qq_line() +
  facet_wrap(restaurant ~., scales = "free")+
  theme(strip.text.y = element_text(angle = 0), aspect.ratio = .6)

```

Out off all Q-Q plots I think that Burger King has distribution closest
to normal because its points follow closer to straight diagonal line
with slightly less deviations comparing with others.

```{r}
burger_king <- sodium_group |>
  filter(restaurant == "Burger King") |>
  ungroup() |>
  as.data.frame()

set.seed(151)
qqnormsim(sample = sodium, data = burger_king)
```

Comparing Burger Kings (BK) data to simulated normal distributions Q-Q
plots confirms that BK's sodium data distribution is is the closest to
normal.

*End of answer.*

8.  Note that some of the normal probability plots for sodium
    distributions seem to have a stepwise pattern. why do you think this
    might be the case?

**Insert your answer here**

*Farhod's answer:*

Let's look at Dairy Queens normal probability plot for sodium
distribution, histogram and at actual values for sodium.

```{r}
ggplot(dairy_queen, aes(sample = sodium)) + 
  geom_line(stat = "qq")
```

```{r}
ggplot(dairy_queen, aes(x = sodium)) +
  geom_histogram(binwidth = 75)
```

```{r}
print(sort(dairy_queen$sodium))
```

I can see stepwise pattern in normal probability plot for sodium values
approximately higher than 1600.

\
The histogram shows significant empty gaps between frequency bins for
sodium values approximately higher than 1600. These empty gaps indicate
a lack of data points within those intervals.\
Also looking at actual Dairy Queens sodium data we can see significant
spread of data between 1610, 2210, 2780 and 3500. \
These findings explain the reasons of stepwise pattern.

*End of answer.*

9.  As you can see, normal probability plots can be used both to assess
    normality and visualize skewness. Make a normal probability plot for
    the total carbohydrates from a restaurant of your choice. Based on
    this normal probability plot, is this variable left skewed,
    symmetric, or right skewed? Use a histogram to confirm your
    findings.

**Insert your answer here**

Farhod's answer:

Let's build a normal probability plot:

```{r}
ggplot(mcdonalds, aes(sample = total_carb)) +
  geom_line(stat = "qq")
```

The plot shows deviations from expected normal probability straight
line. There are significant deviations from the line at the right tail
which suggest that actual data is right-skewed. \
There is also moderate deviation at the left tail, which suggests that
data is not symmetrical.

```{r}
ggplot(mcdonalds, aes(x = total_carb)) +
  geom_histogram()
```

The histogram has outliers in the right tail and confirms that data is
right-skewed with majority of the data points clustered to the left.
Also the shape of the histogram suggests that data not near of being
symmetrical.

------------------------------------------------------------------------
