---
title: "LAB7_DATA607"
author: "Farhod Ibragimov"
date: "2025-03-11"
output: html_document
encoding: UTF-8
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading libraries

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(pdftools)
#library(tabulizer)
library(devtools)
library(dplyr)
library(knitr)
library(htmltools)
library(kableExtra)
library(arrow)
library(rvest)
```

## Converting PDF to dataframe

This code extracts text from PDF file. Then it removes unnecessary lines to keep data only. Finally. the code creates `lab7_inventory_data`dataframe.

```{r}
lab7_pdf <- pdf_text("C:/CUNY_MSDS/DATA607/File_Formats_Assignments.pdf")
cat(lab7_pdf)

txt_lines <- unlist(strsplit(lab7_pdf, "\n"))

head(txt_lines)
#tail(txt_lines)

txt_lines_clean <- txt_lines[-(1:5)]
txt_lines_clean <- txt_lines_clean[-(25:31)]


#print(txt_lines_clean)

lab7_inventory_data <- do.call(rbind, strsplit(txt_lines_clean, ","))
lab7_inventory_data <- as.data.frame(lab7_inventory_data)
colnames(lab7_inventory_data) <- as.character(lab7_inventory_data[1, ])
lab7_inventory_data <- lab7_inventory_data[-1, ]
colnames(lab7_inventory_data)[ncol(lab7_inventory_data)] <- "Description"
head(lab7_inventory_data)
```

Here I do some data cleaning and formatting.

```{r}
lab7_inventory_data$Description[5] <- "Capacity: 20 cu ft"
lab7_inventory_data$Description[8] <- "Capacity: 4.5 cu ft"
lab7_inventory_data$Description[10] <- "Capacity: 5.0 cu ft"
lab7_inventory_data <- lab7_inventory_data[-c(6, 9, 11), ]
head(lab7_inventory_data)
```

## Creating XML data file

This code creates `lab7_inventory.xml`from `lab7_inventory_data`. After that it converts created XML file back into dataframe.

```{r}

library(XML)
lab7_to_xml <- function(df, root_name = "Inventory", item_name = "Item") {
  if (!is.data.frame(df)) stop("Error: Input data must be a dataframe")

  doc <- newXMLDoc()
  root <- newXMLNode(root_name, doc = doc)

  apply(df, 1, function(row) {
    row <- as.list(row)
    row <- lapply(row, as.character)  # Convert all to character

    # Create Item node with attributes for Item Name and ID
    item <- newXMLNode(item_name, attrs = c(Name = row[["Item Name"]], ID = row[["Item ID"]]), parent = root)

    # Add main fields
    newXMLNode("Category", row[["Category"]], parent = item)
    newXMLNode("Brand", row[["Brand"]], parent = item)
    newXMLNode("Price", row[["Price"]], parent = item)

    # Store Variation ID as an attribute
    if (!is.null(row[["Variation ID"]]) && nzchar(row[["Variation ID"]])) {
      variation <- newXMLNode("Variation", attrs = c(ID = row[["Variation ID"]]), parent = item)
      newXMLNode("Details", row[["Variation Details"]], parent = variation)
    } else {
      # Even if Variation ID is missing, create a blank field
      newXMLNode("Variation", attrs = c(ID = ""), parent = item)
    }

    # Add Description
    newXMLNode("Description", row[["Description"]], parent = item)
  })

  return(doc)
}

xml_doc <- lab7_to_xml(lab7_inventory_data)

saveXML(xml_doc, file = "lab7_inventory.xml")

cat(readLines("lab7_inventory.xml"), sep = "\n")
```

```{r}

# reading XML
xml_doc <- xmlParse("lab7_inventory.xml")

# Extract all Item nodes
items <- getNodeSet(xml_doc, "//Item")

# Convert XML to DataFrame (excluding attributes)
lab7_xml_copy <- xmlToDataFrame(nodes = items)

# Extract Variation IDs separately by parsing XML attributes
variation_ids <- sapply(items, function(x) {
  variation_node <- getNodeSet(x, "./Variation")  # Locate Variation node within each Item
  if (length(variation_node) > 0) {
    xmlGetAttr(variation_node[[1]], "ID", default = "")  # Extract "ID" attribute
  } else {
    ""
  }
})

# Add Variation ID column back to DataFrame
lab7_xml_copy$`Variation ID` <- variation_ids

# Print final DataFrame
head(lab7_xml_copy)
```

Pros and cons of XML format:

-   Pros:

    -   Highly structured, extensible, strong schema support.

-   Cons:

    -   Verbose, complex, less common in modern web and data development.

## Creating JSON file format

This code creates `lab7_inventory.json`from dataframe. After that, code converts it back to dataframe.

```{r message=FALSE, warning=FALSE}
library(jsonlite)
lab7_inventory_json <- toJSON(lab7_inventory_data)
write(lab7_inventory_json, "lab7_inventory.json")
cat(readLines("lab7_inventory.json"), sep = "\n")
```

```{r}
json_data <- fromJSON("lab7_inventory.json")
lab7_JSON_copy <- as.data.frame(json_data)
head(lab7_JSON_copy)
```

Pros and cons of JSON format:

-   **Pros:**

    -   Lightweight, human-readable, widely supported.

-   **Cons:**

    -   Less strict schema, limited data types, less efficient for very large datasets.

## Creating HTML format file

Here dataframe is converted to `lab7_inventory.html`. After that, code converts HTML file back into dataframe.

```{r}
html_table <- kable(lab7_inventory_data, format = "html", escape = FALSE)

html_output <- paste0("<html><head><title>Inventory</title></head><body>", 
                      html_table, 
                      "</body></html>")

writeLines(html_output, "lab7_inventory.html")
cat(readLines("lab7_inventory.html"), sep = "\n")

html_file <- read_html("lab7_inventory.html")
tables <- html_file |>
  html_table(fill = TRUE)
lab7_HTML_copy <- tables[[1]]
head(lab7_HTML_copy)
```

Pros and cons of HTML format:

-   **Pros:**

    -   Standard for web pages, widely supported, easy to learn basics.

-   **Cons:**

    -   Primarily for presentation, not ideal for data analysis, can be inconsistent.

## Creating PARQUET format file

This code creates `lab_7inventory.parquet` from dataframe. And after, code converts Parquet file back into dataframe

```{r}
write_parquet(lab7_inventory_data, "lab_7inventory.parquet")
raw_data <- readBin("lab_7inventory.parquet", what = "raw", n = 100)

# Print first 100 bytes
print(raw_data)
```

```{r}

df_parquet <- read_parquet("lab_7inventory.parquet")

head(df_parquet)
```

Pros and cons of PARQUET format:

-   **Pros:**

    -   Columnar storage, highly efficient compression, optimized for big data analytics.

-   **Cons:**

    -   Not human-readable, overhead for small datasets, less suitable for transactional data, slower write speeds.
