---
title: "DATA607_WEEK11"
author: "Farhod Ibragimov"
date: "2025-04-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading libraries

```{r}
library(tidyverse)
library(dplyr)
library(knitr)

```

## Loading movie ratings dataset

```{r}
ratings_df <- read.csv("https://raw.githubusercontent.com/farhodibr/CUNY-SPS-MSDS/refs/heads/main/DATA607/LAB11/movie_reviews%20-%20Form%20Responses%201.csv")

```

This dataset contains movie ratings by users of the survey

```{r}
kable(ratings_df)
```

## Tidying dataset into long format

```{r}
ratings_df_long <- ratings_df |>
  select(-Timestamp) |>
  pivot_longer(
    cols = -Name,
    names_to = "movie",
    values_to = "rating"
  )
kable(head(ratings_df_long, 10))
```

## Global baseline estimate recommender formula

The global baseline estimate recommender uses following formula:

$$
\hat{r}_{ui} = \mu + b_u + b_m
$$

Where:

-   $\mu$ - global average rating

-   $b_u$ - user bias (how a user tends to rate compared to the global average). \
    This calculated by formula:

    $$
    b_u = \frac{1}{N_u} (\sum(r_u - \mu)) 
    $$

    Where: \
    $r_u$ is the movie's rating by user \
    $N_u$ is the number of ratings by the user

-   $b_m$ - movies bias (how a movie is rated compared to the global average)\
    This calculated by formula:

    $$
    b_m = \frac{1}{N_m}(\sum(r_u - \mu - b_u))
    $$ Where: \
    $N_m$ is is the number of ratings for that movie

This code cell calculates $\mu$ (global average rating)

```{r}
mu <- mean(ratings_df_long$rating, na.rm = TRUE)
kable(mu)
```

This code cell creates `user_bias`dataframe, which includes $b_u$ bias for each user:

```{r}
user_bias <- ratings_df_long |>
  group_by(Name) |>
  summarise(bias_user = mean(rating - mu, na.rm = TRUE))
kable(user_bias)
```

This code cell creates `movie_bias` dataset, which include $b_m$ bias for each movie:

```{r}
movie_bias <- ratings_df_long |>
  left_join(user_bias, by =  "Name") |>
  group_by(movie) |>
  summarise(bias_movie = mean(rating - mu - bias_user, na.rm = TRUE))
kable(movie_bias)
```

This code cell creates `predicted_ratings`dataframe which calculates predicted ratings for not rated movies (NAs) by using global baseline estimate recommender formula

```{r}
predicted_ratings <- ratings_df_long |>
  filter(is.na(rating)) |>
  left_join(user_bias, by ="Name") |>
  left_join(movie_bias, by = "movie") |>
  mutate(predicted = round(mu + bias_user + bias_movie))

kable(predicted_ratings)
```
